---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r}
dir.create("./plots")
unzip("./activity.zip") # unzip the file
ACTIV <- read.csv("./activity.csv")
```


## What is mean total number of steps taken per day?

```{r}
require(ggplot2)
# histogram of steps
ggplot2::qplot(steps, data=ACTIV, geom="histogram") + 
    geom_histogram(fill="red", colour="black") + 
    ggtitle("Histogram of Test Subjects' Steps") 

# average steps
paste("The average number of steps is:", mean(ACTIV$steps, na.rm=TRUE))

# median steps
paste("The median number of steps is:", median(ACTIV$steps, na.rm=TRUE))
```


## What is the average daily activity pattern?

```{r}
# calculate a new DF (average steps for a given interval)
ACTIV2 <- aggregate(steps ~ interval, ACTIV, mean, na.rm=TRUE)

# create a line plot of average steps at a given interval
ggplot2::qplot(interval, steps, data=ACTIV2, geom="line") + 
    geom_line(colour="blue") + 
    geom_point(size=1, colour="black") + 
    ggtitle("Average Steps Taken at Each Interval")

dev.copy(png, file="./plots/avg_steps.png")
dev.off()

paste("The interval with the largest steps average is:", 
      ACTIV2$interval[match(max(ACTIV2$steps), ACTIV2$steps)])
```

## Imputing missing values

```{r}
paste("the number of rows with missing data is:", sum(!complete.cases(ACTIV)))
```

- I will fill in missing data by using the mean for the given interval

```{r}
# fill in data using the method mentioned above

# calculate a new DF (average steps for a given interval)
ACTIV2 <- aggregate(steps ~ interval, ACTIV, mean, na.rm=TRUE)
ACTIV_IMPUTE <- ACTIV
# iterate through ACTIV and place in this average value if none exists
for (row in 1:nrow(ACTIV_IMPUTE)) {
    if (is.na(ACTIV_IMPUTE[row, 1])) {
        ACTIV_IMPUTE[row, 1] <- ACTIV2[match(ACTIV_IMPUTE[row, 3], ACTIV2$interval), 2]
    }    
}
rm(list="ACTIV2")

# average steps
paste("The average number of steps in the imputed dataframe is:", 
      mean(ACTIV_IMPUTE$steps, na.rm=TRUE))

# median steps
paste("The median number of steps in the imputed dataframe is:", 
      median(ACTIV_IMPUTE$steps, na.rm=TRUE))

# histogram of steps
ggplot2::qplot(steps, data=ACTIV_IMPUTE, geom="histogram") + 
    geom_histogram(fill="red", colour="black") + 
    ggtitle("Histogram of Test Subjects' Steps (Imputed Data)")

dev.copy(png, file="./plots/hist_steps_impute.png")
dev.off()
```

- The mean and median are equivalent whether or not this imputation occurs

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# create a new DF which is the same as ACTIV but with date coerced to the weekday
ACTIV2 <- transform.data.frame(ACTIV, date=as.factor(weekdays(as.Date(date))))
# classify Monday thru Friday as a weekday, Saturday & Sunday as a weekend
ACTIV2 <- transform.data.frame(ACTIV2, 
                               weekday_ctgry=as.factor(sapply(ACTIV2$date, function(x) if(x %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {"Weekday"} else if(x %in% c("Saturday", "Sunday")) { "Weekend" })))

# calculate a new DF (average steps for a given interval)
ACTIV_PLOT <- aggregate(steps ~ interval + weekday_ctgry, ACTIV2, mean, na.rm=TRUE)

# plot average number of steps taken across a period across Weekday/Weekend
ggplot2::qplot(interval, steps, data=ACTIV_PLOT, facets=.~weekday_ctgry, geom="line") +
    geom_line(colour="darkred") +
    ggtitle("Average Steps at Each Interval for Weekday/Weekend Observations") 

dev.copy(png, file="./plots/panel_steps.png")
dev.off()
```








